Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-10-27 notebooks/expected_and_actual_counts_ratio.Rmd
Add Bonferroni corrected p-values and ensure code works >2 clades

# Create a list of the clades to include in comparisons.
clades <- c("20A","20B")

@@

```{r filter_for_comparisons}
# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
```

To decrease the amount of time it takes to run the contingency tests, add columns to the filtered comparisons data frame with the observed and expected values for clade 1 and clade 2.

```{r contingency_matrix}
# Add observed and expected values for each comparison to filtered comparisons data frame.
{filtered_comparisons$clade1_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed"]
filtered_comparisons$clade1_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"expected"]
filtered_comparisons$clade2_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed"]
filtered_comparisons$clade2_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"expected"]}

#
{filtered_comparisons$clade1_obs_exp <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed_expected"]
filtered_comparisons$clade2_obs_exp <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed_expected"]}
```

@@

# Save and load variable because calculating p-values takes a long time.
'save(p_values, file="fisher_pvalues.Rda")'
load("fisher_pvalues.Rda")
```

@@

```{r p_value_correction}
# False Discovery Rate and Bonferroni Corrections
# NEED TO CORRECT FOR MORE THAN 2 CLADE COMPARISONS
comparison_p_values$p_values_corrected <- comparison_p_values$p_value/nrow(comparison_p_values)
```

@@

```{r volcano_plot}
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
  geom_point()
```

>>>>

# Create a list of the clades to include in comparisons.
clades <- c("20A","20B","20C")

@@

```{r filter_for_comparisons}
# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
# DOUBLE CHECK THIS: NOT CONVINCED IT IS WORKING
filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
```

To decrease the amount of time it takes to run the contingency tests, add columns to the filtered comparisons data frame with the observed counts, expected counts and observed/expected ratios for clade 1 and clade 2.

```{r contingency_matrix}
# Create an ID to maintain the order of rows (and thus comparisons) after merging data frames.
filtered_comparisons$id <- c(1:nrow(filtered_comparisons))

# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
clade1 <- merge(filtered_comparisons[,c("id","clade1")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade1", by.y="nt_mutation_clade")
clade2 <- merge(filtered_comparisons[,c("id","clade2")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade2", by.y="nt_mutation_clade")

# Add observed, expected and observed/expected data to the comparison matrix.
{filtered_comparisons$clade1_observed <- clade1[order(clade1$id),"observed"]
filtered_comparisons$clade1_expected <- clade1[order(clade1$id),"expected"]
filtered_comparisons$clade2_observed <- clade2[order(clade2$id),"observed"]
filtered_comparisons$clade2_expected <- clade2[order(clade2$id),"expected"]
filtered_comparisons$clade1_obs_exp <- clade1[order(clade1$id),"observed_expected"]
filtered_comparisons$clade2_obs_exp <- clade2[order(clade2$id),"observed_expected"]}
```

@@

# Save and load variable because calculating p-values takes a long time.
'save(p_values, file=paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))'
load(paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
```

@@

```{r p_value_correction}
# Identify clades from the two columns in the comparison matrix and combine into one column.
comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,3], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,3], sep="."))

# Count the number of rows for each clade (i.e., number of hypothesis tests).
clade.table <- table(comparison_p_values$clades)

# Add a column for the number of hypothesis tests conducted for each pair of clades.
comparison_p_values$num_tests <- apply(comparison_p_values$clades, 1, FUN=function(c){clade.table[[c[1]]]})

# Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
comparison_p_values$p_values_corrected <- comparison_p_values$p_value/comparison_p_values$num_tests
```

@@

```{r volcano_plot}
# Only plot data for the two clades being compared.
plot_data <- comparison_p_values[which(comparison_p_values$clades=="20B.20C"),]
ggplot(plot_data, aes(x=fold_change, y=-log10(p_value)))+
  geom_point()
```
